Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-09-29 docs/MEE_submission/manuscript_MEE.tex
Updated the MS with some small changes

In many different contexts, researchers make use of data presented in primary literature. In the fields of ecology and evolution (E\&E), this data is most commonly used for comparative and meta-analyses. Meta-analysis in particular is a rapidly expanding methodology in E\&E, not only in terms of the number of meta-analyses (in plant ecology alone the yearly number of published meta-analyses doubled from 2006 to 2012 (20-40); \cite{Koricheva2014}), but also in terms of their size (a recent meta-analysis, for example, included 6440 effect sizes from 175 publications; \cite{Noble2018}). The integrative nature of these studies makes them extremely important to the progression of E\&E \citep{Gurevitch2018}. These techniques rely heavily on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures, including both standalone programs and \proglang{R} packages (reviewed in Table \ref{tab:comparison}), these tools do not cater to the general needs of meta-analysts for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). 
%
 First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}.
%
 Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later, typically using different software. % Given the ubiquity of the R platform in E\&E, and that the most popular meta-analysis software in E\&E (e.g., metafor and MCMCglmm) are available as R packages, it is highly likely to be used for some (if not all) stages of the research synthesis process. It would therefore be very beneficial to have comprehensive, and robust, digitisation capabilities in R that provides users with lots of flexibility, and makes the process of figure extraction more transparent and easier to reproduce.
%
 Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images and export the resulting digitised data into individual files one-by-one. These data often subsequently need to be imported and edited using different software. 
%
 Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently, a large amount of additional data manipulation is required, that is different across plots types. For example, in E\&E data are commonly presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented. Histograms and box plots are also frequently used in E\&E to presented data, and whilst their downsream calculations are even more laborious, there are few (if any; see Table \ref{tab:comparison}) tools to extract data from these plot types.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Although line extraction is not generally necessary in meta-analysis and comparative analysis, outside of these fields researchers may need to extract parameters of a line from a figure. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

>>>>

In many different contexts, researchers make use of data presented in primary literature. In the fields of ecology and evolution (E\&E), these data are most commonly used for comparative and meta-analyses. The use of meta-analysis in E\&E in particular, is rapidly growing, not only in terms of the number of meta-analyses (in plant ecology alone the yearly number of published meta-analyses doubled from 2006 to 2012 (20-40); \cite{Koricheva2014}), but also in terms of their size (a recent meta-analysis, for example, included 6440 effect sizes from 175 publications; \cite{Noble2018}). Meta-analyses are extremely important in providing a means of quantitatively synthesizing experimental and/or observational studies to evaluate empirical support for fundamental theory in E\&E \citep{Gurevitch2018}. These techniques rely heavily on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures. For example, in some cases up to 42\% of the papers used have one or more figures for data extraction \cite{Noble2018}. These data often need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures, including both standalone programs and \proglang{R} packages (reviewed in Table \ref{tab:comparison}), these tools do not cater to the general needs of meta-analysts for four main reasons (here we focus on meta-analysis, although many points apply to extraction for comparative analysis). 
%
 First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}.
%
 Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later, typically using different software. % Given the ubiquity of the R platform in E\&E, and that the most popular meta-analysis software in E\&E (e.g., metafor and MCMCglmm) are available as R packages, it is highly likely to be used for some (if not all) stages of the research synthesis process. It would therefore be very beneficial to have comprehensive, and robust, digitisation capabilities in R that provides users with lots of flexibility, and makes the process of figure extraction more transparent and easier to reproduce.
%
 Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images and export the resulting digitised data into individual files one-by-one. These data often subsequently need to be imported and edited using different software. 
%
 Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently, a large amount of additional data manipulation is required, that is different across plots types. For example, in E\&E data are commonly presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented. Histograms and box plots are also frequently used in E\&E to presented data, and whilst their downstream calculations are even more laborious, there are few (if any; see Table \ref{tab:comparison}) tools to extract data from these plot types.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Although line extraction is not generally necessary in comparative and meta-analysis, outside of these fields researchers may need to extract parameters of a line from a figure. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.
